---
title: "STAT545A Individual Mini Data Analysis"
author: "Sara Kowalski"
date: "October 9, 2021"
output: github_document
---



## Table of Contents
1. Introduction

2. Install and load packages
    + 2.1 Install **datateachr**
    + 2.2 Load **datateachr** and **tidyverse**
    
3. Task 1: Choosing a dataset
    + 3.1 Choose 4 datasets
    + 3.2 Explore the 4 datasets
        + 3.2.1 *vancouver_trees* 
        + 3.2.2 *cancer_sample* 
        + 3.2.3 *flow_sample*
        + 3.2.4 *steam_games*
    + 3.3 Narrowing it down to 2 datasets
    + 3.4 Final decision: choose a dataset
    
4. Task 2: Exploring the dataset
    + 4.1 Diving deeper into the data 
        + 4.1.1 Exercise 1
        + 4.1.2 Exercise 2
        + 4.1.3 Exercise 3
        + 4.1.4 Exercise 4
        
5. Task 3: Research questions
    + 5.1 Question #1 
    + 5.2 Question #2 
    + 5.3 Question #3 
    + 5.4 Question #4 

## 1  Introduction
Hello! The following document contains the work required to complete Milestone 1 of the Mini Data Analysis Project in STAT545A. Here I will be using R, and more specifically the tidyverse packages within R, to explore and visualize the data within the *vancouver_trees* dataset. 

Much of the inspiration for this script comes from [Heads or Tails](https://www.kaggle.com/headsortails/tidy-titarnic) and from the [instructions](https://stat545.stat.ubc.ca/mini-project/mini-project-1/) provided for the project. 

## 2  Install and Load Packages 
### 2.1 Install **datateachr**

```{r}
## The following code must be commented out in order to knit the rmd file. If necessary, unncomment the
## following code to install the packages. 
## install.packages("devtools")
## devtools::install_github("UBC-MDS/datateachr")

```

### 2.2 Load **datateachr** and **tidyverse**

```{r}
## Load dataset package - datateachr
library(datateachr)

## Load tidyverse - a collection of R packages 
library(tidyverse) 
```

## 3  Task 1: Choosing a dataset
### 3.1 Choose 4 datasets
Within the **datateachr** package by Hayley Boyce and Jordan Bourak, there are 7 available datasets. I have choosen 4 that interest me the most to explore further:

**My 4 Choices**

1.  *vancouver_trees*
2.  *cancer_sample*
3.  *flow_sample*
4.  *steam_games*

### 3.2 Explore the 4 datasets
To narrow down the datasets and make a decision on which one to choose, I have explored each dataset using the *class*, *summary*, and *glimpse* functions

* *Class* function: provides the class(es) the specified object called is from 
* *Summary* function: provides a description for each column within the specified dataset 
* *Glimpse* function: produces the observations and variables contained in the specified dataset, along with a brief look at the oultine of the dataset 

#### 3.2.1 *vancouver_trees*
Exploring the *vancouver_trees* dataset that is courtesy of The City of Vancouver's Open Data Portal. 

```{r}
## explore the *vancouver_trees* dataset using the class, summary, and glimpse functions
class(vancouver_trees)
summary(vancouver_trees)
glimpse(vancouver_trees)
```
Attributes of the *vancouver_trees* dataset:

* 146 611 rows 
* 20 columns:
    + Numeric columns (7)
    + Character columns (12)
    + Date column (1)
* column **cultivar_name** contains NA values
* classes:
    + tbl_df
    + tbl
    + data.frame

#### 3.2.2 *cancer_sample*
Exploring the *cancer_sample* dataset that is courtesy of UCI Machine Learning Repository.

```{r}
## explore the *cancer_sample* dataset using the class, summary, and glimpse functions
class(cancer_sample)
summary(cancer_sample)
glimpse(cancer_sample)
```
Attributes of the *cancer_sample* dataset:

* 569 rows
* 32 columns:
    + Numeric columns (31)
    + Character columns (1)
* classes: 
    + spec_tbl_df
    + tbl_df
    + tbl
    + data.frame

#### 3.2.3 *flow_sample*
Exploring the *flow_sample* dataset that is courtesy of The Government of Canada's Historical Hydrometric Database.

```{r}
## explore the *flow_sample* dataset using the class, summary, and glimpse functions
class(flow_sample)
summary(flow_sample)
glimpse(flow_sample)
```
Attributes of the *flow_sample* dataset:

* 218 rows
* 7 columns:
    + Numeric columns (4)
    + Character columns (3)
* columns **month**, **day**, **flow**, and **sym** contain NA values
* classes:
    + tbl_df
    + tbl
    + data.frame

#### 3.2.4 *steam_games*
Exploring the *steam_games* dataset that is courtesy of Kaggle.

```{r}
## explore the *steam_games* dataset using the class, summary, and glimpse functions
class(steam_games)
##summary(steam_games)
##glimpse(steam_games)
```
Attributes of the *steam_games* dataset:

* 40 833 rows
* 21 columns:
    + Numeric columns (4)
    + Character columns (17)
* columns **achievements**, **mature_content**, **original_price**, and **discount_price** contain NA values
* classes:
    + spec_tbl_df
    + tbl_df
    + tbl
    + data.frame

### 3.3 Narrowing it down to 2 datasets
After a brief exploration of all 4 datasets, I have narrowed my decision down to 2 datasets:

    1.  *vancouver_trees*
    2.  *cancer_sample*

**Why these datasets?** Both of these data sets contain the most numeric variables (*vancouver_trees* (7); *cancer_sample* (31)) out of all 4 datasets I have explored. This will allow me to have more freedom when exploring and graphing the data as there are more options for manipulating and displaying numeric data compared to character data (i.e. mathematical manipulations, distribution graphs, etc). As well, the high number of columns in both datasets allows for me to ask more in-depth, complex questions about the data and test different combinations of columns. Overall, both of these datasets contain a vast amount of information that gives me the opportunity to develop intriguing and thought provoking research questions that I will answer in Milestones 2 and 3 of this Mini Data Analysis Project. And it does not hurt that I am very passionate about hiking and the outdoors and am currently doing my Master's Thesis in a cancer biology lab, making both of these datasets very up my alley!     

### 3.4 Final decision: choose a dataset
In choosing my final dataset, I have formulated 1 research question for each dataset that I would be interested in exploring:

1. *vancouver_trees*: How have the number of trees planted within each neighbourhood changed over time?

2. *cancer_sample*: Does the radius of the sample impact the patient's diagnosis (Malignant (M) vs. Benign (B))? 

**Final Decision: *vancouver_trees* **

I have chosen the *vancouver_trees* dataset to work with throughout this Mini Data Analysis Project because I found it fitting seeing as I am currently living in Vancouver and love everything to do with the outdoors and exploring nature. Plus, I am currently surrounded by all things cancer for my masters, so exploring the *vancouver_trees* dataset will be a nice breath of fresh air :)

## 4  Task 2: Exploring the dataset
### 4.1 Diving deeper into the data
To dive deeper and further explore the *vancouver_trees* dataset, I have completed 4 exercises. This will aid in my understanding of the dataset and help the development of more meaningful questions to be answered in future milestones.

#### 4.1.1 Exercise 1
I have choosen to create a density plot that contains the distribution of the diameters of the trees contained in the dataset. Since the *vancouver_trees* dataset has a very high number of rows (i.e. lots of trees), I am able to easily visualize all of my data in one plot and see if there is a large or wide distribution of tree diameters in the data or if the distribution is narrow. This will then allow me to decided whether I want to further filter or subset my data down to look at specific groups of trees (e.g. trees with a large diameter) which will also shrink the amount of data within my dataset making it easier to work with and explore certain relationships between other variables. 
```{r}
## Plotting the distribution of the diameter of the trees 
## use the ggplot function for visualizations
ggplot(vancouver_trees, aes(diameter)) + ## histograms only require that x is a numeric vector 
  geom_histogram(bins = 50) ## bins specifies the number of bars shown in the plot
```

#### 4.1.2 Exercise 2
I have decided to filter the observations in the *vancouver_trees* dataset so I am able to better explore the tree data I am most interested in. The *vancouver_trees* dataset is quite a large dataset containing 146611 rows. This means there are a lot of different listed species, genus, and common names for the trees within the data set, and that makes it extremely difficult to graph any of those variables or their relationship with the other variables within the dataset. By filtering down the data (in this case to only contain trees with the genus Ulmus), I am better able to visualze the tree data as it is no longer squished or 'running' off the screen. 
```{r}
## Filtering the data using the dplyr filter function 
## create a variable that holds a subset of the data and give it a descriptive name
ulmusGenus <- vancouver_trees %>%
## filter the data so only trees with the genus name ULMUS remain
  filter(genus_name == "ULMUS")
## visualize your new subsetted dataset
print(ulmusGenus)
```

#### 4.1.3 Exercise 3
Using a ridge plot to look at the relationship between the diameter of the trees for each species within the Ulmus genus. I want to explore the relationship between two variables in the dataset so I have decided to look at the relationship between a categorical variable (species) and a numerical variable (diameter) to see how they relate. This way I can see the distribution of tree diameters within a species and determine if certain tree species within the Ulmus genus have larger diameters than others or if there is a large distribution of tree diameters within a species, etc.  

**I will be using the filtered data, ulmusGenus, obtained from exercise 2 as the full *vancouver_trees* dataset is too large to plot the distribution graphs for the diameters of the trees for all the genus' listed. If I were to use the whole *vancouver_trees* dataset, the graph would be extremely long and very crowded, and it would be nearly impossible to make meaningful observations about the data**
```{r}
## use ggplot to visualize the data
## call the dataset you wish to use and then specify which variables within that dataset you would like to graph
## chose to graph the categorical variable on the y axis to clean up the presentation of the data and prevent crowding 
ggplot(ulmusGenus, aes(diameter, species_name)) +
  ggridges::geom_density_ridges(aes(fill = species_name))
```

#### 4.1.4 Exercise 4
I have chosen to create a new variable called species_diameter_mean which will hold the mean tree diameter for each species listed in the dataset. I could also create a new varibale called genus_diameter_mean which would hold the mean tree diameter for each genus in the dataset and give a more global or zoomed out view of the data. Creating these new variables will let me explore new relationships will other variables more easily (less crowded graphs) and from a new angle.
```{r}
## use the dplyr function mutate to create a new variable called species_diameter_mean in the
## *vancouver_trees* dataset
## call the dataset
vancouver_trees %>%
## group by species_name to group together the species in the dataset 
  group_by(species_name) %>%
## use mutate to add the new variable species_diameter_mean by calculating the mean diameter of the trees
## within each species and assigning the output to the new specified variable. This works because mutate
## honors the group function.
  mutate(species_diameter_mean = mean(diameter))

## use the dplyr function mutate to create a new variable called genus_diameter_mean in the
## *vancouver_trees* dataset
## call the dataset
vancouver_trees %>%
## group by genus_name to group together the genus' in the dataset 
  group_by(genus_name) %>%
## use mutate to add the new variable genus_diameter_mean by calculating the mean diameter of the trees
## within each genus and assigning the output to the new specified variable. This works because mutate
## honors the group function.
  mutate(genus_diameter_mean = mean(diameter))

```

## 5  Task 3: Research questions
Now that I have explored and become familiar with the *vancouver_trees* dataset, I am ready to generate my 4 research questions. 

### 5.1 Question 1
What is the distribution for tree species (from a specified genus (e.g. Ulmus)) within Vancouver neighbourhoods? Are certain tree species favoured over others in specific neighbourhoods? 

### 5.2 Question 2
How does the amount of trees planted change over time? Can go more in depth and look at how this changes within the different Vancouver neighbourhoods listed in the dataset. Or can look at how the number of trees planted changes for each genus or species over time. Can also look at a specific genus or species and see how the amount of trees planted changes over time within a specific neighbourhood or all the neighbourhoods in Vancouver.   

### 5.3 Question 3
Within each genus, what is the distribution of the diameter of the trees for each species? Which species have a wider diameter distribution and which ones have a more narrow distribution?  

### 5.4 Question 4
Which tree species are found planted on curbs? Does this vary by neighbourhood? Does the diameter of the tree determine whether it is planted on a curb (Y) or not (N) (i.e. large diameter trees would not be planted on curbs and smaller diameter trees would be)? Will have to group the trees by genus in order to pair down the dataset because for this question, it is too vast to look at in its entirety. 
