---
title: "STAT545A Individual Mini Data Analysis-Milestone 2"
author: "Sara Kowalski"
date: "October 19, 2021"
output: github_document
---



## Table of Contents
1. Introduction

2. Load packages
    + 2.1 Load **datateachr** and **tidyverse**
    
3. Task 1: Process and summarize *vancouver_trees*
    + 3.1 My 4 research questions
    + 3.2 Summarizing and graphing 
        + 3.2.1 Question 1
            + 3.2.1.1 Summarizing
            + 3.2.1.2 Graphing
        + 3.2.2 Question 2
            + 3.2.2.1 Summarizing
            + 3.2.2.2 Graphing
        + 3.2.3 Question 3
            + 3.2.3.1 Summarizing
            + 3.2.3.2 Graphing
        + 3.2.4 Question 4
            + 3.2.4.1 Summarizing
            + 3.2.4.2 Graphing
    + 3.3 Reflect on results 
    
4. Task 2: Tidy *vancouver_trees* dataset
    + 4.1 Is *vancouver_trees* tidy? 
    + 4.2 untidy, then tidy back *vancouver_trees*
    + 4.3 Choose 2 research questions
    + 4.4. Choose a version of *vancouver_trees*


## 1  Introduction
Hello! The following document contains the work required to complete Milestone 2 of the Mini Data Analysis Project in STAT545A. Here I will be using R, and more specifically the tidyverse packages within R, to summarize and visualize/graph the data within the *vancouver_trees* dataset. By the end of this document I will have chosen 2 research questions to continue to explore in Milestone 3 and generated a version of my data that will help me to best answer those questions.  

Much of the inspiration for this script comes from [Heads or Tails](https://www.kaggle.com/headsortails/tidy-titarnic) and from the [instructions](https://stat545.stat.ubc.ca/mini-project/mini-project-2/) provided for the project. 

## 2  Load Packages 
### 2.1 Load **datateachr** and **tidyverse**

```{r}
## Load dataset package - datateachr
library(datateachr)

## Load tidyverse - a collection of R packages used for data analysis
library(tidyverse) 
```

## 3  Task 1: Process and summarize *vancouver_trees*
### 3.1 My 4 research questions

* Question 1:
What is the distribution for tree species (from a specified genus (e.g. Ulmus)) within Vancouver neighbourhoods? Are certain tree species favored over others in specific neighbourhoods? 

* Question 2:
How does the amount of trees planted change over time? Can go more in depth and look at how this changes within the different Vancouver neighbourhoods listed in the dataset. Or can look at how the number of trees planted changes for each genus or species over time. Can also look at a specific genus or species and see how the amount of trees planted changes over time within a specific neighbourhood or all the neighbourhoods in Vancouver.   

* Question 3:
Within each genus, what is the distribution of the diameter of the trees for each species? Which species have a wider diameter distribution and which ones have a more narrow distribution?  

* Question 4:
Which tree species are found planted on curbs? Does this vary by neighbourhood? Does the diameter of the tree determine whether it is planted on a curb (Y) or not (N) (i.e. large diameter trees would not be planted on curbs and smaller diameter trees would be)? Will have to group the trees by genus in order to pair down the dataset because for this question, it is too vast to look at in its entirety. 

### 3.2 Summarizing and graphing
To explore the *vancouver_trees* dataset more in depth, I will use various data manipulation techniques to help me complete a summarization and graphing activity for each research question listed above. 

#### 3.2.1 Question 1
##### 3.2.1.1 Summarizing
For this question it would be useful to know how many tree species there are for a given tree genus and how many Vancouver neighbourhoods there are in the dataset. Knowing these counts, I can determine how much data I am working with and how big my graphs will be. This will help me to determine the best way of presenting the data and inform my decision on whether to subset the data further or not. With that being said, I will use the *distinct* function from the *dplyr* package to determine the amount of observations for the *neighbourhood_name* variable and the *species_name* variable in the *vancouver_trees* dataset. 

```{r}
## first call the data set that will be used 
vancouver_trees %>%
## next call the distinct function and input the variable from the dataset you wish to know the count for 
  distinct(neighbourhood_name)

## repeat as necessary for any other variables within the dataset. 
vancouver_trees %>%
## I want to know the amount of distinct species within a particular tree genus so I will first filter
## the dataset to only contain trees from that genus 
  filter(genus_name == "ULMUS") %>%
  distinct(species_name)
```

Based on the tibble outputs, there are **22 distinct neighbourhoods** within the *vancouver_trees* dataset and **7 distinct species** within the Ulmus genus in the dataset. I think it would also be interesting to take this analysis one step further and see the count for the amount of trees in each of the 7 disinct species in the Ulmus genus. Since each row in the dataset is a tree, I can compute this using the function *count*. 

```{r}
## call the dataset
vancouver_trees %>%
## filter the data so it only contains trees with the genus ULMUS 
  filter(genus_name == "ULMUS") %>%
## use the count function and input the variable you want the disinct values for
  count(species_name)
```

Based on this further analysis, I can see that the species **AMERICANA** has the most counts (2252) while **CARPINIFOLIA** has the least (17). 

**In conclusion** : I have successfully determined the counts for the amount of neighbourhoods within the dataset and the number of species within the Ulmus genus while taking it a step further and determining the amount of trees per species within the Ulmus genus. This summarization exercise gives me a better idea at the amount of data I am working with and how it could be further subsetted. For instance, I may want to group the neighbourhoods into quadrants (north, west, east, and south) and that way I can look at larger number of genus and species of trees within each quadrant instead of each neighbourhood. 

##### 3.2.1.2 Graphing
To start answering this question, I am going to plot a bar graph that will look at the amount of trees for each species in the **STYRAX** genus. I am also going to colour the bar graph by neighbourhood so that it will become a stacked bar graph holding more information about where trees are located and how many species of each tree are loocated in each Vancouver neighbourhood. Further I am going to plot another bar graph with the neighbourhood as the x-axis and the colour of the bars as the species to compare and see which plot is cleaner and easier to interpret.  

```{r}
## define a new place to hold the data to be graphed and call the dataset to be used
styraxGenus <- vancouver_trees %>%
## filter that data so it only contains trees with the STYRAX genus
  filter(genus_name == "STYRAX") %>%
## group the data by species_name in order to get the number of trees in each species
  group_by(species_name) %>%
## use the mutate function to add a new column that contains the count of trees for each distinct species 
## in the STYRAX genus
  mutate(numTrees = n())

## plot the data using ggplot, set x axis to species_name
ggplot(styraxGenus, aes(species_name, fill = neighbourhood_name)) +
## set the graph to a bar graph
  geom_bar() +
## define the x and y axis labels 
  xlab("Species") +
  ylab("Number of Trees")

## plot the data using ggplot, set x axis to neighbourhood_name
ggplot(styraxGenus, aes(neighbourhood_name, fill = species_name)) +
## set the graph to a bar graph
  geom_bar() +
## define the x and y axis labels 
  xlab("Neighbourhood") +
  ylab("Number of Trees") +
## declutter the bottom axis by switching x and y 
  coord_flip()
```

**In conclusion** : I have successfully plotted the graphs to compare the amount of trees planted per species in the **STYRAX** genus within each vancouver neighbourhood in the dataset. I found that plotting the data with the neighbourhood on the y axis and the fill set to species that the data was a lot easier to interpret and looked much cleaner. This exercise has helped me to successfully answer my research question one. To take this analysis further, I could try adding a graph for each genus or for a select few genus' by facet wrapping.  

#### 3.2.2 Question 2
##### 3.2.2.1 Summarizing
For this research question, instead of looking at how trees planted changes over time, I could look at how trees planted changes depending on the 4 seasons: summer, fall, winter, and spring. More specifically, I can look and see if there is a specific season in which trees are preferably planted and see if this varies by neighbourhood or type of tree (genus or species). A useful summarization activity for this question would be to create a new categorical variable called *season* from the date_planted variable to determine which season each tree was planted in. For the purposes of this activity, I will define the 4 seasons as follows:

* summer = July (7) - September (9)
* fall = October (10) - December (12)
* winter = January (1) - March (3)
* spring =  April (4) - June (6)

In order to do this I must first add a variable to the dataset called *month* that will hold numerical data corresponding to the month that each tree was planted based on the trees *date_planted* value. To ensure reproducibility of the code and ease of comprehension, I will create a new dataset called *numericMonth* which will store this added variable *months* along with the original data from the *vancouver_trees* dataset. 

```{r}
## define where the data created will be stored and which dataset will be used to create the new data
numericMonth <- vancouver_trees %>%
## add the new variable with the mutate function
## to access the month of each date use the months.Date function
## to further change the month into a numerical variable use the as.numeric and as.factor functions
  mutate(month = as.numeric(as.factor(months.Date(date_planted))))
## use the glimpse function to confirm the addition of the new variable - month
glimpse(numericMonth)
```

Now I can change the numeric variable *month* into a categorical variable *season* that holds the 4 seasons (groups): summer, fall, winter, and spring. I will store this in a new dataset called *season_van_trees_planted*. 

```{r}
## define where the data created will be stored and which dataset will be used to create the new data
season_van_trees_planted <- numericMonth %>%
## use the mutate function to add the new variable season 
## use the cut function to create the categorical variable season with 4 groups (summer, fall, winter,
## spring) from the numeric variable month 
## winter(1-3), spring(4-6), summer(7-9), fall(10-12)
  mutate(season = cut(month, breaks = c(1,3,6,9,12), labels = c("winter", "spring", "summer", "fall")))

## use the glimpse function to confirm the addition of the new variable - season 
glimpse(season_van_trees_planted)
```

**In conclusion** I have successfully created a new categorical variable called *season* with 4 groups: winter, spring, summer, and fall, out of transformed date data. In further analysis, I would likely filter out all the NA values from the data to clean it up or graph it with NA as it's own group. This summarization exercise will help to graph the data when I explore the relationship between species of trees and the month they were planted in.

##### 3.2.2.2 Graphing
I am now able to graph this data using the new summarized variable *season* I just created. I have decided to graph a basic bar graph to look at the amount of **ULMUS** trees that have been planted in each neighbourhood throughout the 4 seasons. 

```{r}
## define a new place to hold the data to be graphed 
ulmusGenus_season_planted <- season_van_trees_planted %>%
## Filter to only trees with the ULMUS genus 
  filter(genus_name == "ULMUS")

## take the new filtered data set ulmusGenus_season_planted and input that into ggplot and use geom_bar
## to create a bar graph with the x axis set to season 
## this will plot the amount of Ulmus trees planted in each season 
## set fill to neighbourhood_name 
ggplot(ulmusGenus_season_planted, aes(season, fill = neighbourhood_name)) +
  geom_bar(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~neighbourhood_name) +
  theme_minimal()
```

**In conclusion** I have successfully created a bar graph to visualize the amount of **ULMUS** trees that have been planted in each neighbourhood per season. Since there are so many neighbourhoods in the dataset it makes it hard to visualize each individual graph I have created. Moving forward it would be wise to either narrow my question to specific neighbourhoods within the dataset or group the neighbourhoods into 1 of 4 quadrants (north, east, south, west) and look at the amount of **ULMUS** trees that have been planted in each quadrant of the city per season. 

#### 3.2.3 Question 3
##### 3.2.3.1 Summarizing
For this research question, it may be helpful to look at a handful of summary statistics for the diameter of the trees across the species of a specific genus. For this exercise I will be calculating the **range**, **mean**, **median**, and **standard deviation** of tree diameters for each species within the Ulmus genus. Computing these summary statistics will help me understand how the distribution of tree diameter might vary between different species, and could help me to ask deeper questions to develop a more complex analysis about the relationship between tree diameter and species. 

I will compute these statistics by utilizing the *summarize* function within the *dplyr* package.

```{r}
## call the dataset
vancouver_trees %>%
## filter the data for the Ulmus genus 
  filter(genus_name == "ULMUS") %>%
## group the data by species 
  group_by(species_name) %>%
## calculate the range, mean, median, and standard deviation of tree diameter for each species with the
## summarize function
  summarize(avg_treeDiameter = mean(diameter), range_treeDiameter = range(diameter), median_treeDiameter = median(diameter), sd_treeDiameter = sd(diameter))
```

**In conclusion** I have successfully calculated summary statistics about the diameters of the trees within the species of the Ulmus genus. Based on the results, I can see that the species **AMERICANA** has the largest range of diameters with the minimum being 1.25 and the maximum being 144.00 and thus if I were to plot this data, I would predict that the **AMERICANA** species would have the widest distribution of tree diameters. It would take more calculations to determine this with certainty, however, these initial calculations give me a good start at answering my original research question. 

##### 3.2.3.2 Graphing
I am now going to complete a graphing exercise for this research question. I am going to look at the distribution of the diameter of the trees for each species within the **ULMUS**, **STYRAX**, and **PYRUS** genus. My code for this chunk of data was inspired by the code written for question 4.3 within the [STAT545A worksheet 3](https://github.com/UBC-STAT/stat545.stat.ubc.ca/blob/master/content/worksheets/worksheet_a03.ipynb).  

```{r}
# call the dataset 
vancouver_trees %>% 
## filter for the specific tree genus' needed
  filter(genus_name == c("ULMUS", "STYRAX", "PYRUS")) %>% 
## plot the data using ggplot
## x = diameter, y = genus_name
  ggplot(aes(diameter, genus_name)) +
## clean up the plot by using ggridges to look at the distribution of diameters in each species
## set the fill of each ridge to be the species_name and set the alpha transparency to 1/3 so each 
## ridge will be clearly visible
  ggridges::geom_density_ridges(aes(fill = species_name), alpha = 1/3) +
## rename the y axis to clean up the graph
  ylab("Genus") +
  theme_minimal() +
## set the colour scheme 
  scale_fill_discrete("")
```

**In conclusion** I have successfully created a ridges distribution graph to visualize the distribution of tree diameters across tree species of three different tree genus'. Although I did use alpha transparency to attempt to better see each distribution, in the future it would be useful to maybe facet the data by genus so that each graph can be seen on its own and hopefully be easier to read. This exercise does not do the best job at answering my original question as it is difficult to see the tree diameter distribution for each species and is only looking at three genus' within the dataset whereas I would like to look at all the genus'. Given the size of the dataset, I may want to narrow down my research question to look at only a portion of the genus' given. That way my data analysis will be much cleaner and easier to interpret. 

#### 3.2.4 Question 4
##### 3.2.4.1 Summarizing
This research question would also benefit from the summarization exercise performed above for question 3. However, to look at this in a different light and further explore my data, I will perform similar calculations as the ones above with a slight twist. Instead of grouping the trees by species and then calculating diameter summary statistic, I will group the trees by whether (Y) or not (N) they have been planted on a curb. This will let me look at the types of diameters trees on curbs have vs. those not planted on curbs. For this exercise I will calculate the **range**, **mean**, **IQR**, and **standard deviation**. The trees will be filtered by genus, specifically looking at the **ACER** genus.  

```{r}
## call the dataset
vancouver_trees %>%
## filter the data for the Acer genus 
  filter(genus_name == "ACER") %>%
## group the data by curb 
  group_by(curb) %>%
## calculate the range, mean, IQR, and standard deviation of tree diameter for each species with the
## summarize function
  summarize(avg_treeDiameter = mean(diameter), range_treeDiameter = range(diameter), IQR_treeDiameter = IQR(diameter), sd_treeDiameter = sd(diameter))
```

**In conclusion** I have successfully calculated summary statistics about the diameters of the trees planted on curbs (Y) vs. those not planted on curbs (N) within the **Acer** genus. Based on the results, I can see that trees planted on curbs have a slightly lower average (mean) tree diameter compared to those not planted on curbs. However, it is also evident that trees planted on curbs have a much larger range of tree diameters compared to those not planted on curbs. This is a good start to answering my full research question and could help me to subset my data for further analysis. 

##### 3.2.4.2 Graphing 
To begin answering part of my original research question, I am going to create a jitter plot that is going to look at just the species within the **ULMUS** genus and see if they have been planted on a curb or not. 

```{r}
## define a new place to hold the data to be graphed and call the dataset needed 
ulmusGenus_curb_planted <- vancouver_trees %>%
## filter for only trees with the genus ULMUS
  filter(genus_name == "ULMUS")

## use ggplot to graph the data
## define the dataset being used
## set x axis to species_name and y axis to curb
ggplot(ulmusGenus_curb_planted, aes(species_name, curb)) +
## specify the type of plot by using geom_jitter to create a jitter plot 
## set alpha transparency to 1/10 
  geom_jitter( alpha = 1/10) +
## re-label the x axis to clean up the graph  
  xlab("Species") +
## set theme to visually customize the graph  
  theme_light()
```

**In conclusion** I have successfully calculated created a jitter plot to visualize the amount of trees in each species of the **ULMUS** genus that have been planted on curbs (Y) or not (N). I made the analysis of the data easier by setting the alpha transparency to 1/10. Based on the results, I can see that the **AMERICANA** species has the most trees planted on curbs compared to the other species, with **GLABRA** having the second most, and **PUMILA** having the third most. To further sophisticate this graph and delve deeper into my analysis for this question, in the future I am going to try incorporating the neighbourhoods the trees have been planted in by maybe setting the colour of each point to equal a corresponding neighbourhood.  

### 3.3 Reflect on results
Based on all the results from the summarizing and graphing exercises, I am a bit closer to answering my original research questions. Question 1 could essentially be answered with the work that I completed above (summarizing and graphing); however, I want to make the analysis more complex and look at graphing each tree genus. Question 2 is a bit more complicated and would likely require more data rearrangement and potentially looking at the city by quadrants instead of individual neighbourhoods. Question 3 is close to being answered, but I need to figure out a way to more cleanly/clearly display the ridge graphs and also include more genus types along with maybe finding a way to include some summary statistics on the graph such as the mean, max, and min values. Finally question 4 would also require a lot more data rearrangement in order to incorporate the neighbourhoods into the analysis. Out of all the questions so far, I think question 1 has yielded some of the most interesting results mainly because the graph is very easy to interpret and so patterns and distributions are much easier to comprehend.  

## 4  Task 2: Tidy *vancouver_trees* dataset
### 4.1 Is *vancouver_trees* tidy?
It is important when you are setting up to answer any research question given a dataset that you determine whether your data is in tidy form or not. Looking through the *vancouver_trees* dataset, I have determined that the data is in **tidy form**. Each row is an observation, each column is a variable, and each cell holds a value. This can be demonstrated nicely by using the *genus_name* and *species_name* columns. Currently, within the dataset, the genus and species names of each tree is separated into 2 columns. This allows for tidy data as each cell is holding one value, a species name or a genus name. If however, the genus and species name were grouped together into the same column called *genus_species* the cells of this column would be holding two values which could make the analysis of the data more difficult depending on the question asked. Another easy example is the *civic_number* and *std_street* columns. These are currently separated and hold a number and a character value, respectively. In comparison, if this street data were to be grouped into one column, the number values from *civic_number* would no longer exist as they would have to be turned into character variables in order to squish the street data into one column that would hold the street address the tree is planted on. Again this could limit the analysis depending on the question asked.  
```{r}
## quick look at the dataset
glimpse(vancouver_trees)
```

### 4.2 untidy, then tidy back *vancouver_trees*
To see how the data would look if it were untidy and to get an understanding for how difficult some of the tasks would be if I were to use an untidy dataset to answer my research questions, I am going to untidy the *vancouver_trees* dataset by widening it. To do this I am going to pull out the data from the *common_name* column and make each distinct value its own column. This values for the new columns will be the genus and species names which will effectively remove the *genus_name* and *species_name* columns from the dataset. 

```{r}
## define the new dataset and call the dataset you are wanting to untidy
untidy_vancouver_trees <- vancouver_trees %>%
## use the pivot_wider function and specify which columns to remove/keep from the dataset
## specify where the names for the new columns will come from and where the values for the new columns will
## come from
    pivot_wider(id_cols = c(-common_name, -species_name), 
                names_from = common_name,
                values_from = species_name)
glimpse(untidy_vancouver_trees)
```

At first glance this new untidy dataset is very, very messy and large now. This would be extremely hard to work with when answering the research questions I have posed. Given the large amount of data within this dataset, I will also demonstrate untidying the data using a subset of the original dataset that only looks at the **ULMUS** genus

```{r}
## define the new subest of data and the dataset being used to create it 
untidy_ulmusGenus <- vancouver_trees %>%
  filter(genus_name == "ULMUS") %>%
## use the pivot function and specify which columns to remove/keep from the dataset
## specify where the names for the new columns will come from and where the values for the new columns will
## come from
  pivot_wider(id_cols = c(-common_name, -species_name), 
              names_from = common_name, 
              values_from = species_name)
glimpse(untidy_ulmusGenus)
```

Now I will tidy back the *untidy_ulmusGenus* dataset by using the pivot_longer function! (I am not going to tidy back the *untidy_vancouver_trees* dataset as it is very large and will unnecessarily take up space)

```{r}
tidy_ulmusGenus <- untidy_ulmusGenus %>% 
  pivot_longer(cols = -c(tree_id:latitude), 
               names_to  = "common_name", 
               values_to = "species_name")
glimpse(tidy_ulmusGenus)
```

### 4.3 Choose 2 research questions
Based on the analysis conducted in this Milestone, I have decided to pursue questions 1 and 3 in the future Milestone 3 activities. 

* Question 1:
What is the distribution for tree species (from a specified genus (e.g. Ulmus)) within Vancouver neighbourhoods? Are certain tree species favored over others in specific neighbourhoods? 

* Question 3:
Within each genus, what is the distribution of the diameter of the trees for each species? Which species have a wider diameter distribution and which ones have a more narrow distribution?  

I have chosen these two questions because **(a)** since I have moved to Vancouver, I have lived in several different areas (Kitsilano, Mount Pleasant, Point Grey, Kerrisdale, etc), and I thought it would be interesting to see the distribution of trees throughout the neighbourhoods I have lived in (Question 1). **(b)** I think it would be interesting to see how the diameter varies between the genus of trees and between the species of trees within a genus (Question 3) and based on the preliminary graphing I did for this question, I think it would be cool to try and display this data with a ridge plot. Finally, **(c)** I will likely be using different functions to manipulate and plot the data in order to formulate the answers to either question which will allow me to gain experience with multiple different functions within *dyplr* and *tidyverse*.  

### 4.4. Choose a version of *vancouver_trees*
To make the analysis of questions 1 and 3 easier, I am now going to rearrange the *vancouver_trees* dataset into a version I feel will be more appropriate to work with. 

```{r}
## give a name to the new version and call the dataset you are going to be rearranging
vancouver_trees_milestone3 <- vancouver_trees %>%
## keep only the columns need for the analysis of questions 1 and 3 with the select function
  select(genus_name, species_name, neighbourhood_name, diameter)
## this will be the main dataset, I will now split it into two datasets that will be tailored to answering
## questions 1 and 3

## dataset for question 1
## give a name to the new version and call the dataset you are going to rearrange
vancouver_trees_milestone3_q1 <- vancouver_trees_milestone3 %>%
## filter for the desired neighbourhoods
  filter(neighbourhood_name %in% c("MOUNT PLEASANT", "OAKRIDGE", "KITSILANO", "KERRISDALE", "WEST POINT GREY")) %>%
## group by genus since I want to look at each species within each genus  
  group_by(genus_name) %>%
## then use summarize and n_distinct to count the number of distinct species within each genus. 
  mutate(n_species = n_distinct(species_name))

## dataset for question 3
## give a name to the new version and call the dataset you are going to rearrange
vancouver_trees_milestone3_q3 <- vancouver_trees_milestone3 %>%
## group by genus and species
  group_by(genus_name, species_name) %>%
## add summary statistics that may be useful for graphing 
  mutate(avg_treeDiameter = mean(diameter), median_treeDiameter = median(diameter), sd_treeDiameter = sd(diameter), max_treeDiameter = max(diameter), min_treeDiameter = min(diameter))
```
  
