---
title: "STAT545A Individual Mini Data Analysis-Milestone 3"
author: "Sara Kowalski"
date: "October 19, 2021"
output: github_document
---



## Table of Contents
1. Introduction

2. Load packages
    + 2.1 Load **datateachr** and **tidyverse**
    + 2.2 My research questions
    
3. Task 1: Special data types
    + 3.1 Reordering factors 
    + 3.2 Modifying time 

4. Task 2: Modelling
    + 4.1 Research question and variable of interest
    + 4.2 Hypothesis testing
        + 4.2.1 P-value 

5. Task 3: Reading and writing data
    + 5.1 Write a csv file
    + 5.2 Write a R binary file


## 1  Introduction
Hello! The following document contains the work required to complete Milestone 3 of the Mini Data Analysis Project in STAT545A.  

Much of the inspiration for this script comes from [Heads or Tails](https://www.kaggle.com/headsortails/tidy-titarnic) and from the [instructions](https://stat545.stat.ubc.ca/mini-project/mini-project-3/) provided for the project.

## 2  Load Packages 
### 2.1 Load **datateachr** and **tidyverse**

```{r}
## Load dataset package - datateachr
library(datateachr)

## Load tidyverse - a collection of R packages used for data analysis
library(tidyverse) 
```

### 2.2 My Research Questions

1. Question 1:
What is the distribution for tree species (from a specified genus (e.g. Ulmus)) within Vancouver neighbourhoods? Are certain tree species favored over others in specific neighbourhoods? 

2. Question 2:
Within each genus, what is the distribution of the diameter of the trees for each species? Which species have a wider diameter distribution and which ones have a more narrow distribution?

## 3  Task 1: Special data types
### 3.1 Reordering factors

For this task I have decided to take a plot I produced in milestone 2 which involved plotting across the three groups: *neighbourhood_name*, *number of trees*, and *species_name*, and produce a new plot with the neighbourhoods reordered in ascending order (i.e. the neighbourhood with the least amount of trees on the far left with it increasing up to the neighbourhood with the most amount of trees on the far right). I will complete this task using the tools in the **forcats** package which was designed for working with and reordering factors. 

Here is the code and a visual of the original plot I created in Milestone 2 which looked at how many trees in the **STYRAX** genus were planted in each Vancouver neighbourhood. The graph was further subsetted to show how many of the different types of **STYRAX** species were planted in each neighbourhood by colouring the bars based on the *species_name*.  

```{r}
## define a new place to hold the data to be graphed and call the dataset to be used
styraxGenus <- vancouver_trees %>%
## filter that data so it only contains trees with the STYRAX genus
  filter(genus_name == "STYRAX") %>%
## group the data by species_name in order to get the number of trees in each species
  group_by(species_name) %>%
## use the mutate function to add a new column that contains the count of trees for each distinct species 
## in the STYRAX genus
  mutate(numTrees = n())

## plot the data using ggplot, set x axis to neighbourhood_name
ggplot(styraxGenus, aes(neighbourhood_name, fill = species_name)) +
## set the graph to a bar graph
  geom_bar() +
## define the x and y axis labels 
  xlab("Neighbourhood") +
  ylab("Number of Trees") +
## declutter the bottom axis by switching x and y 
  coord_flip()
```

To more clearly see how the number of trees planted changes in each neighbourhood, I am going to reorder the neighbourhoodds so they appear in ascending order from left to right based on the number of **STYRAX** trees planted. 

```{r}



```